---
title: "GEE for multilevel data"
output:
  html_document:
    df_print: paged
    keep_md: true
---

This is a breif intorduction using GEE on simulation data as an example.

## generating simulation data

First, some parameters requires to be set.
```{r, message=FALSE}
require(MASS)
source("functions_simul.R")  # loading required functions

n_smp <- 200  # sample size
n_lv <- 8  # number of levels (including base level)
# n_rep <- 1000  # times of repeats
corr_coef <- 0.2  # correlation coefficient for simulation data

## grenrating real values of thetas randomly
random_seed <- 123
set.seed(random_seed)
theta_r <- rand_theta(n_lv)
print(theta_r)

sdata <- simul_data(theta_r, n_smp, n_lv, corr = corr_coef)
str(sdata)
```

`sdata` contains 2 parts:

* `V` is the data matrix for simulation($n_{sample}{\times}n_{level}$)  
* `theta` records ${\theta}s$  


note: `theta_r` is a list of 4 types of ${\theta}s$, so can also be determined manually as follow:
```{r, eval=FALSE}
## manually setting real values of thetas
theta_r <- list(zeta = 2.3, 
                gamma0 = c(0.0, -0.4,  1.5, 1.8, 
                           -1.8,  0.1, 1.6, 0.2), 
                eta = 2.4, 
                gamma1 = c(1.0, 0.9, 1.9, 0.9, 
                           1.4, 1.1, 0.2, 1.8))
```


## an example for modeling on simulation data with gee

`gee_solve` estimates ${\theta}s$ by Newton's method with moment estimating for initialization

* input:  
    + `V`: a matrix of raw data
    + `iter_times`: maximum times of iteration
    + `eps`: threshold for stopping iteration
* output:  
    + `theta_hat`: estimates of ${\theta}s$
    + `sd_hat`: estimates of standard deviates
    + `sigma_hat`: estimates of covariances
    + `theta_iters`: a list of estimates of ${\theta}s$ for each iteration

```{r}
M_raw <- sdata$V
n_lv <- ncol(M_raw)
source("functions_GEE.R")
esti_gee <- gee_solve(M_raw, iter_times = 1000, eps = 1e-6)
str(esti_gee, max.level = 2)
```

`get_adj` transforms raw data to adjusted data with estimates of ${\theta}s$
```{r}
M_gee <- get_adj(M_raw, esti_gee$theta_hat)
```

`get_p` helps to get p-values
```{r}
p_raw <- get_p(m = M_raw, 
               test_f = friedman.test, 
               adjust = "BH", 
               n = n_lv - 1)
p_gee <- get_p(m = M_gee, 
               test_f = friedman.test, 
               adjust = "BH", 
               n = n_lv - 1)
sapply(list(raw = M_raw, adj = M_gee), 
       get_p, 
       test_f = friedman.test, 
       adjust = "BH", 
       n = n_lv - 1)
```

## simulating for 1000 times
```{r, results='hold'}
n_smp <- 200  # sample size
n_lv <- 8  # number of levels (including base level)
n_rep <- 1000  # times of repeats
corr_coef <- 0.5  # correlation coefficient for simulation data

## grenrating real values of thetas randomly
random_seed <- 123
set.seed(random_seed)
theta_r <- rand_theta(n_lv)

theta_real <- 
  list(zeta = theta_r$zeta, 
       gamma0 = theta_r$gamma0[-1], 
       eta = theta_r$eta, 
       gamma1 = theta_r$gamma1[-1])
theta_real <- unlist(theta_real)

source("functions_GEE.R")

t0 <- proc.time()
theta_all_iterations <- c()
sd_all_iterations <- c()
for(i in 1:n_rep){
  set.seed(i)
  M_raw <- simul_data(theta_r, n_smp, n_lv, corr = corr_coef)
  M_raw <- M_raw$V
  esti_gee <- gee_solve(M_raw, eps = 1e-6)
  # M_gee <- get_adj(M_raw, esti_gee$theta_hat)
  theta_esti <- unlist(esti_gee$theta_hat)
  sd_esti <- esti_gee$sd_hat
  theta_all_iterations <- rbind(theta_all_iterations, theta_esti)
  sd_all_iterations <- rbind(sd_all_iterations, sd_esti)
}
t1 <- proc.time()
cat("time consuming:\n")
print(t1 - t0)
cat("\n")

theta_esti <- colMeans(theta_all_iterations)
sd_emp <- sqrt(sapply(as.data.frame(theta_all_iterations), var))
sd_esti <- colMeans(sd_all_iterations)
cbind(real = theta_real, 
      esti_gee = theta_esti, 
      sd_emp = sd_emp, 
      sd_gee = sd_esti)
```

a comparison to linear regression
```{r, results='hold'}
lm_solve <- function(df){
  df <- sdata$V
  lm.outs <- list()
  for(i in 2:dim(df)[2])
  {
    colnames(df) <- NULL
    df <- as.data.frame(df)
    colnames(df)[1] <- "y"
    colnames(df)[i] <- "x"
    lm.out <- lm(y ~ x, df)
    lm.outs$zeta <- NA
    lm.outs$gamma0 <- c(lm.outs$gamma0, lm.out$coefficients[[1]])
    lm.outs$eta <- NA
    lm.outs$gamma1 <- c(lm.outs$gamma1, lm.out$coefficients[[2]])
  }
  lm.outs
}

t0 <- proc.time()
lm_all_iterations <- c()
for(i in 1:n_rep){
  set.seed(i)
  M_raw <- simul_data(theta_r, n_smp, n_lv, corr = corr_coef)
  M_raw <- M_raw$V
  esti_lm <- unlist(lm_solve(M_raw))
  lm_all_iterations <- rbind(lm_all_iterations, esti_lm)
}
t1 <- proc.time()
cat("time consuming:\n")
print(t1 - t0)
cat("\n")

theta_lm <- colMeans(lm_all_iterations)
cbind(real = theta_real, 
      esti_gee = theta_esti, 
      esti_lm = theta_lm)
```

